---
title: "R crash course: A quick introduction to R"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and  \n
        Santiago Perez-Hoyos"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Copenhagen"
    colortheme: "dolphin"
    fonttheme: "structurebold"
  slide_level: 2
footer: "R for Data Science"

---

```{r setLicense, child = 'license.Rmd'}
```

# A Crash Course in R

## Outline

- Why R
    + R basics
    + How does one work with R and Rstudio
- Getting Started
    + A primer of data import
    + Variables and data types
    + Functions, Packages and more stuff
- Working with data
    + Selecting, Filtering and ordering datasets
    + A primer of statistics and plots
    + R Notebooks and RMarkdown

## Motivation

- We (you) all work with data, most of the time and often we need to do "things" with those data.

    - I have three lists of genes and I would like to see which genes they have in common (or which  ones appear only in one list).
    - We have received the data from that lab but I only want to work with a subset of the samples.
    - Is it possible to repeat that plot changing the line colors, the font size etc?
    - I have some scripts tu re-run an analysis but I don't know how to start

- These, and many other things can be done with a basic knowledge of R.

## [What is R?](https://www.r-project.org/about.html)

- R is a _language and environment_ for statistical computing and graphics.

- R provides a wide variety of statistical and graphical techniques, and is highly extensible.

- It can be used fro simple tasks to highly complex reproducible projects.

- It compiles and runs on a wide variety of UNIX platforms and similar systems Windows and MacOS.

<!-- ## R PRO's (why you are here!) -->

<!-- - The system is  -->
<!--     + free (as in _free beer_) -->
<!--     + It's platform independent -->
<!--     + It is constantly improving (2 new versions/year) -->

<!-- - It is a statistical tool -->
<!--     + Implements almost every statistical method that exists -->
<!--     + Great graphics (Examples) -->
<!--     + Simple reporting tools -->
<!--     + Also state-of-the-art in Bioinformatics through the [Bioconductor Project](http://bioconductor.org). -->

<!-- - Programming language -->
<!--     + Easy to automate repetitive tasks (Example_1.1) -->
<!--     + Possibility to create user friendly web interfaces  with a moderate effort. (Examples) -->

<!-- ## R CON's -->

<!-- - R is mainly used issuing commands from a console -->
<!--     + less user friendly than almost any other statistical tool you may know. -->

<!-- - Constantly having new versions may affect our projects -->

<!-- - Not necessarily the best language nor suitable for every existing task -->

## How is R used

- Different ways to use R, but the best trade-off simplicity-efficiency is provided by Rstudio

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/RStudio.png}
\end{figure}

## Exercise

- Get to know R. Visit the R-project page and see what can be found there.

- If you haven't done it before, download and install R and Rstudio in your computer

- Open R studio. Look at the panels and figgure out what can we do at each window.


# Using R

## Commands, Objects and Functions

- Shortly, using R consists of 
    + Working with *objects* using *commands* and *functions*

\begin{figure}
\includegraphics[width=0.85\linewidth]{images/basicConcepts.png}
\end{figure}

## Variables and data types

- Data managed in R ...
    + is stored as *variables*
- Variables can be of distinct types
    + Numerical
        - numeric (13.7)
        - int (3)
    + Character
        - "R is cute"
    + Factors
        - A,B,C,D
        - WT, Mut

## R packages

- R can be used for many different types of data processing and analysis from distinct fields, besides statistics such as Ecology, Omics Sciences, Psychology etc.
- All these capabilities are not present from the begining because most of them will never be used by most users.
- Instead, thay can be added when needed by 
    (i) installing and 
    (ii) loading the appropriate packages.

## Installing and loading packages
 
We want to analyze some data using cox proportional hazards model.
```{r eval=FALSE}
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
```
`Error in coxph(Surv(time, status) ~ sex, data = lung) : could not find function "coxph"`

We need to install and load the package before we can use it.
```{r eval=FALSE}
install.packages("survival")
library(survival)
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
```

## Bioconductor

<!-- \begin{figure} -->
<!-- \includegraphics[width=0.85\linewidth]{images/basicConcepts.png} -->
<!-- \end{figure} -->

- Packages analyse all kinds of Genomic data (>800)
- Compulsory documentation (*vignettes*) for each package
- 6-month release cycle
- [Course Materials](http://bioconductor.org/help/course-materials/)
- [Example data](http://bioconductor.org/packages/release/BiocViews.html#___ExperimentData) and [workflows](http://bioconductor.org/help/workflows/)
- Common, re-usable framework and functionality
- [Available Support](https://support.bioconductor.org/)
    + Often you will be able to interact with the package maintainers / developers and other power-users of the project software

## The `tidyverse`

- The tidyverse is an opinionated collection of R packages designed for data science. 
- All packages share an underlying design philosophy, grammar, and data structures. 
- The complete tidyverse collection can be installed with:
```{r eval=FALSE}
install.packages("tidyverse")
```
- [https://www.tidyverse.org/](https://www.tidyverse.org/)


# Getting data into R

## Importing data with Rstudio

- The easiest way to get data into R is to click on the `Ìmport Datasets` button.
- Alternatively R code can be written using functions from `Base R` or the `tidyverse`
    + `Base R` functions start with `read.`: `read.table`, `read.csv`
    + `tidyverse` functions start with `read_`: `read_delim`, `read_csv` or `read_excel`

## Reading Excel or csv files

- Files can be read from any location, let it be a physical support or a web site.
- To read files from disk be sure to indicate their location.
- Alternatively the default working directory can be set to the folder where the file is located.

## Reading Excel or csv files (continued)

- Assume files `TIO2+PTYR-human-MSS+MSIvsPD.XLSX` has been downloaded to your working directory

- Start setting the default directory to the folder where you have saved the file.
    + `Session --> Set Working directory --> To source file location...`

- Import the `TIO2+PTYR-human-MSS+MSIvsPD.XLSX` with the default options

## Reading Excel or csv files (continued)

- Code generated for reading the files can be reused any time changing the file name if needed.

```{r, eval=FALSE}
# Read Excel file
library(readxl)
otherData <- read_excel("otherFiles")
```


## Interlude: Summarizing data

- Once a dataset is available it is easy to "have a look at it"

```{r eval=FALSE}
head(phosphoprotData)
str(phosphoprotData)
summary (phosphoprotData)
```

# Dynamic output with Rmarkdown

## Reproducible research with R notebooks

- R and Rstudio are strongly involved in promoting [reproducibility](https://en.wikipedia.org/wiki/Reproducibility) and [reproducible research](https://en.wikipedia.org/wiki/Reproducibility#Reproducible_research).
- This is implemented in **R notebooks**
- A notebook combines
    + Natural language text, e.g. describing what we are doing in our own words.
    + R code with the instructions needed to do the data management or the analysis.
    + The output of the analysis

## Creating Notebooks

- A notebook can be created in Rstudio with 
    + `File --> New File --> R Notebook`
- The notebook contains example text and code so it is straightforwoard to adapt it to your analysis.
- To produce an html file with text, code and output:
    - Press the button "Preview"
    - Or Select "Knitr to Html"

# Resources and exercises

## Introductory materials

The web is full of all types of materials about R

Below there are a couple of brief introductions:

- [A short introduction to R]( https://stats.idre.ucla.edu/r/seminars/short-introduction-to-r/)

- [Getting started with R](https://github.com/saghirb/Getting-Started-in-R)

## Exercise

- Select a dataset with which you wish to work along the course.
- Read it into R
    + How many variables are there in it
    + What are their types
- Try to summarize it briefly
- Create an R notebook to encapsulate all your steps and share it with somebody.