---
title: "Session 8 - Categorical data Analyisis. Exercises"
output:
  html_notebook:
    code_folding: show
---

# Exercise 1

- Select `diabetes.xls` datasets 
    
 - Read the dataset into R and check that the categorical variables you are interested (mort, tabac, ecg) in are converted into factors.
 
 - Confirm the conversion by summarizing the variables
 
```{r}
library(readxl)
diabetes <- read_excel("datasets/diabetes.xls")
str(diabetes)

diabetes$mort <- as.factor(diabetes$mort)
diabetes$tabac <- as.factor(diabetes$tabac)
diabetes$ecg <- as.factor(diabetes$ecg)

```
 
 

  
  
# Exercise 2

- With the diabetes dataset repeat the crosstabulation done above using 
  - Two categorical variables
  - Variable "mort" and a newly created variable "bmi30" created by properly categorizing variable bmi.
  
```{r}
library(dplyr)
diabetes <- diabetes %>% mutate(bmi30 = bmi >30 )

mt <- addmargins(table(diabetes$mort, diabetes$bmi30))
mt
prop.table(mt)
```


# Exercise 3

- In the diabetes dataset.
  
  - Test the hypothesis that the proportion of patients with `bmi30` is higher than 40%
      - In the global population of the study
      - Only in patients with 'mort' equal "Muerto"
      
```{r}
prop.test(table(diabetes$bmi30), p = 0.4)

diab_mort <- diabetes %>% filter(mort == "Muerto")
prop.test(table(diab_mort$bmi30), p = 0.4)
```
  

<!-- - Select a sample of size 100 and repeat the test. How do the results change? -->

<!-- - What sample size should we have taken so that th precision of the confidence intervals would have been at most 3% with a probability of 95%? -->

 
