---
title: "Session 5. Introduction to statistical inference"
output:
  pdf_document: default
  html_notebook:
    code_folding: hide
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# Example 1

- Read the Osteoporosis dataset and turn factors into variables automatically with Rbase function `read.delim`
- Take a sample of size 100 from the original file. Call it 'osteo100' and work with this file from now on.
- Compute the mean value of the variable containing bone density values `BUA`
- Split the computation between all subgroups from variable `classific` and variable `menop`.
- Compute the percentage of menopausic women from variable `menop`.

```{r}
library(dplyr)
# Read data
osteoporosis <- read.delim2("datasets/osteoporosis.csv", stringsAsFactors=TRUE)
# Take subsample
osteo100 <- sample_n(osteoporosis, 100)
# mean bone density
buaMean <- mean(osteo100$bua)
print(buaMean)
# Mean bone density ny groups
osteo100 %>% 
  group_by(menop) %>% 
  summarize(m = mean(bua))
# Proportion of menop women (Proportion  is a mean of 0-1 values)
mean(ifelse(osteo100$menop=="SI",1,0))
```


# Exercise 1

- Read the diabetes dataset. Convert characters into factors before continuing.
- Provide an estimate of 
    - The distribution of a numerical variable.
    - a proportion of at least one categorical variable and 
    - the mean value of at least one numerical variable.
- Could you have used different estimators?
- How would you decide?

First we read data and recode character values into factors.

```{r}
library(readxl)
library(dplyr)
library(magrittr)
diabetes <- read_excel("datasets/diabetes.xls")
sapply(diabetes, class)
diabetes_factor <- diabetes %>%
  mutate_if(sapply(diabetes, is.character), as.factor) %>%
  select (-numpacie)
sapply(diabetes_factor, class)
```
Next provide a quick summary of each variable
```{r}
summary(diabetes_factor)
```
Plotting all variables with an instruction is a bit tricky. May be easier to plot separately numerical and categorical variables.

```{r}
library(ggplot2)
library(tidyr)
library(purrr)

diabetes %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```

Proceed similarly with categorical variables

```{r}
diabetes %>%
  keep(is.character) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_bar()
```

You may notice -or not- that the dataset has some outlier values.

Before removing them consider estimating the mean nvalue of SBP and DBP with distinct estimators

```{r}
with(diabetes_factor, {
    print("DBP")
    show(summary(dbp))
    print("SBP")
    show(summary(sbp))
  }
  )
```
What is prefereable to estimate the mean SBP or DBP?



## Example 2. Computing Confidence Intervals with R (2)

```{r}
t.test(osteo100[["bua"]])
```
---

## Example 2 . Computing Confidence Intervals with R (3)

```{r}
cntMenop <- table(osteo100[["menop"]])["SI"]
ssize <- length(osteo100[["menop"]])
prop.test (x=cntMenop, n=ssize)
```

# Exercise 2.1 Computing Confidence intervals

- Read the file "osteoporosis.csv" into a dataset and call it "osteoporosis"

- Compute confidence intervals for the BUA mean and for the percentage of menopausic women with __all the individuals in the dataset__.

- Compare these confidence intervals with those that you obtained in example 2. How do they differ?

**La solución en el tema 6**

# Exercise 2.2 Computing Confidence intervals

- Read the diabetes dataset. Convert characters into factors before continuing.



```{r diabetes}
library(readxl)
library(dplyr)
library(magrittr)
diabetes <- read_excel("datasets/diabetes.xls")
sapply(diabetes, class)
diabetes_factor <- diabetes %>%
  mutate_if(sapply(diabetes, is.character), as.factor) %>%
  select (-numpacie)
sapply(diabetes_factor, class)

```


- Provide a confidence interval for:
  - a proportion of at least one categorical variable and 
  - the mean value of at least one numerical variable.
    
```{r cisdiab}
cnt <- table(diabetes[["mort"]])["Muerto"]
ssize <- length(diabetes[["mort"]])
prop.test (x=cnt, n=ssize)

t.test(diabetes[["edat"]])
```
    

- How would you find alternative approaches to compute these confidence intervals?

  - An option is to apply formulas directly, calculating from:
  $$
  \bar{x}\pm t_{\alpha/2}\frac{s}{\sqrt{n}}
  $$

- Why would you want to do such a thing?

```{r}
mostra<-diabetes[["edat"]]

m<-mean(mostra)  # Calculate mean
 
sd<-sd(mostra)  # Calculate standard deviation 
se<-sd/sqrt(length(mostra)) # Calculate standard Error
li<- m-qt(.975,length(mostra)-1)*se # Calculate 95% CI lower bound

ls<- m+qt(.975,length(mostra)-1)*se # Calculate 95%CI upper bound


cat("Mean=",m,"\n")
cat("Standard deviation=",sd,"\n")
cat("Standard error=",se,"\n")
cat("95% Confidence interval=(",li,";",ls,")","\n")
```

- An *approximate* confidence interval for proportions can also be computed using a normal approximation such as:

$$\hat{p}\pm z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$ 


``` {r eje2_1}
cnt <- table(diabetes[["mort"]])["Muerto"]
ssize <- length(diabetes[["mort"]])
p<-cnt/ssize
n<-ssize 
z<-qnorm(.975)
ee<-sqrt((p*(1-p))/n)
lowerli<- p-z*ee
upperli<- p+z*ee

cat("95% confidence interval for  ", p ,"=(",lowerli,";",upperli,")","\n")

```

# Example 3. Sample size calculation

- Using the osteoporosis dataset, assume that the standard deviation is a good aproximation to $\sigma$.

- Find the sample size needed to achieve a margin of error equal to 5 with a $95\%$ confidence interval.

- This can be computed with distinct packages.
- An option is the `sample.size.mean`function from the `samplingbook` package.

- TYpe `? sample.size.mean` to learn about it 


```{r}
library(samplingbook)
osteoporosis <- read.delim2("datasets/osteoporosis.csv", stringsAsFactors=TRUE)
sdbua<-sd(osteoporosis$bua)
merror<-5
sample.size.mean(merror,sdbua)

```


# Exercise 3. Sample size calculation

- The sample size formula for proportions is: 

$$
n= \frac{\hat{p}(1-\hat{p})z_{1-\alpha/2}^2} {\Delta^2},
$$
where $\Delta$ is the margin error, that is the maximum expected difference between the true value and its estimation, that one expects to have with a probability of, at least $1-\alpha$.

- Write a function to compute the sample size for proportions in the worst case (p=q=0.5) or assuming $p$ is known.

- Using a $50\%$ planned proportion estimate, find the sample size needed to achieve $5%$ margin of error for a survey at $95%$ confidence level.

- How would this result change if we are told that a pilot study suggests that $p=10\%$?
```{r pssize}

alpha<-1-.95
z<-qnorm(1-alpha/2)
merror<-0.05
p<-0.5 # Worst proportion 
nsample<- (p*(1-p) * z^2)/0.05^2
cat("Sample size for 95% CI and 5% margin error s ",round(nsample))

p<-.1 # Worst proportion 
nsample<- (p*(1-p) * z^2)/0.05^2
cat("Sample size for 95% CI and 10% margin error s ",round(nsample))

```




```{r}
t.test(osteoporosis$bua,mu=72.5
       )
```

