---
title: "Session 8 - Categorical data Analyisis. Exercises"
output:
  html_notebook:
    code_folding: hide

---

# Exercise 1

- Select one of the datasets that you have worked with during the course
 
  - `diabetes.xls`
  - `osteoporosis.csv` (in this case select a random sample of 100 observations)
  - `demora.xls`
    
 - Read the dataset into R and check that the categorical variables you are interested in are converted into factors.
 
 - Confirm the conversion by summarizing the variables
 
```{r}
library(dplyr)
semilla<- 18101998
set.seed(semilla)
numMuestras <- 100
muestras <- sample( osteoporosis$registro, numMuestras)
osteoData <- dplyr::filter(osteoporosis, registro  %in%  muestras)
```
 
 
 
# Exercise 2
 
- Use the diabetes.sav file and import it into R with the "Import from SPSS" feature.

  - What is the class of the "MORT" variable.
  
  - Turn it into one factor so that it has the same levels as when you read it using `read.csv`
  
  
# Exercise 3

- With the osteoporosis dataset repeat the crosstabulation done above using:
  - Two categorical variables
  - Variable "MENOP" and a newly created variable "catBUA" created by properly categorizing variable BUA.
  
- Take a random sample of 100 observations or use the file `osteo100.txt`

```{r}
osteo100 <- read.delim("datasets/osteo100.txt", stringsAsFactors=TRUE)
osteoOrderd <- osteo100[order(osteo100$bua),]
osteo100$buaCat <- cut (osteo100$bua, c(0,50, 75, 100)) #https://www.r-bloggers.com/2012/09/from-continuous-to-categorical/
table(osteo100$buaCat)
with(osteo100, table(buaCat, clasific))
```

# Exercise 4

- In the osteoporosis dataset.
  
  - Test the hypothesis that the proportion of women with `osteoporosis` is higher than 7%
      - In the global population of the study
      - Only in women with osteoporosis
      
- Select a sample of size 100 and repeat the test. How do the results change?
  
- What sample size should we have taken so that th precision of the confidence intervals would have been at most 3% with a probability of 95%?

 

# Exercise 5

- Use the osteoporosis dataset to study if it can be detected an association between the variables `menop` and `classific` in the osteoporosis dataset.

- Do not start with a test but with an appropriate summarization and visualization!

