---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

# Data managements with dplyr

## filter

```{r}
require(readxl)
diab <-read_excel("diabetes.xls")

head(diab, n = 4)
summary(diab)

diab <- diab %>% mutate_if(is.character, as.factor)
summary(diabetes)

require(tidyverse)
#require(dplyr)
diab_filt <- filter(diab, tabac=="No fumador", edat>=50) # and & - or |

head(diab_filt, n = 4)

summary(diab_filt)
#diab$tabac <- as.factor(diab$tabac) ##**Warning!** by default alphabetic order is used when creating factor levels.

rm(diabetes) #Drop dataset from env
```

## select

```{r}
diab_small <- select(diab, mort, edat, tabac, sbp)
head(diab_small, n = 4)

diab_small_consec <- select(diab, mort:bmi) # Select consecutive variables
diab_small_noconsec <- select(diab, !(mort:bmi)) # Non-select consecutive variables

```

## sort

```{r}
diab_sort <-arrange(diab,desc(mort), edat)
head(diab_sort, n = 4)
```

## mutate

```{r}
diab_mut <- mutate(diab,edatcat =cut(edat, breaks =c(0,40,50,60,70,120)),highbmi = bmi> mean(bmi))
tail(diab_mut, n = 4)

summary(diab_mut)

table(diab_mut$edatcat, diab_mut$highbmi)

diab_mut_2 <- mutate(diab,edatcat =cut(edat, breaks =c(0,40,50,60,70,120),
                include.lowest=TRUE,
                labels=c("0-40","41-50","51-60", "61-70", "71-120")))
```

## E1

Find all individual that:

1.  Had a sbp higher than 160 (filter())

```{r}
diab_sbp <- filter(diab, sbp>160)
summary(diab_sbp)
```

2.  Had a sbp higher than 160 or tabac was 'Fumador'

```{r}
diab_sbp_tabac <- filter(diab, sbp>160 | tabac=='Fumador')
summary(diab_sbp_tabac)
```

    -   What happens if you include the name of a variable multiple times in a select() call?

```{r}
diab_mselect <- select(diab, sbp, sbp, dbp)
summary(diab_mselect);names(diab_mselect)
```

3.  Sort individual to find the most 'tempsviu'. (arrange() )

```{r}
diab_ltempsviu <- arrange(diab, tempsviu)
head(diab_ltempsviu, 4)
```


# The pipe operator

```{r}
diab40 <-filter(diab, mort=="Vivo"&edat<40)
diab40_small <-select(diab40, edat, dbp)
head(diab40_small,n = 4)
```

```{r}
diab40_small <- diab %>%
  filter(mort=="Vivo"&edat<40) %>%
  select(edat, dbp)
head(diab40_small,n = 4)
```

## E2

Replicate the last exercice using 'pipes':
```{r}
df <- filter(diab,sbp > 160 | tabac == "Fumador")

dfs <- select(df, tempsviu ,bmi,sbp,sbp)

dfsa <- arrange(dfs, desc(tempsviu))
```

```{r}
diab_pipes <- diab %>% filter(sbp > 160 | tabac == "Fumador")  %>%
  select(tempsviu ,bmi,sbp,sbp) %>% 
  arrange(desc(tempsviu))
```



# Merging datasets

```{r}
band_members
band_instruments
```

```{r}
bind_rows(band_members, band_instruments)

## PROBLEMES!
bind_cols(band_members, band_instruments)
```

## mutating

```{r}
band_members %>% inner_join(band_instruments, by = "name")
```

```{r}
band_members %>% left_join(band_instruments)
```

##E3
What happens if you run these lines?
```{r}
band_members %>% right_join(band_instruments)

band_members %>% full_join(band_instruments)
```



#More
```{r}
diab %>% group_by(mort) %>% summarise(mean(sbp))
```

