---
title: "Session 5. Introduction to statistical inference"
output:
  html_document:
    df_print: paged
---

# Example 1

- Read the Osteoporosis dataset and turn factors into variables automatically with Rbase function `read.delim`
- Take a sample of size 100 from the original file. Call it 'osteo100' and work with this file from now on.
- Compute the mean value of the variable containing bone density values `BUA`
- Split the computation between all subgroups from variable `classific` and variable `menop`
- Compute the percentage of menopausic women from variable `menop`

```{r}
library(dplyr)
# Read data
osteoporosis <- read.delim2("datasets/osteoporosis.csv", stringsAsFactors=TRUE)
# Take subsample
osteo100 <- sample_n(osteoporosis, 100)
# mean bone density
buaMean <- mean(osteo100$bua)
print(buaMean)
# Mean bone density ny groups
osteo100 %>% 
  group_by(menop) %>% 
  summarize(m = mean(bua))
# Proportion of menop women (Proportion  is a mean of 0-1 values)
mean(ifelse(osteo100$menop=="SI",1,0))
```


# Exercise 1

- Read the diabetes dataset. Convert characters into factors before continuing.
- Provide an estimate of 
    - The distribution of a numerical variable.
    - a proportion of at least one categorical variable and 
    - the mean value of at least one numerical variable.
- Could you have used different estimators?
- How would you decide?

First we read data and recode character values into factors.

```{r}
library(readxl)
library(dplyr)
library(magrittr)
diabetes <- read_excel("datasets/diabetes.xls")
sapply(diabetes, class)
diabetes_factor <- diabetes %>%
  mutate_if(sapply(diabetes, is.character), as.factor) %>%
  select (-numpacie)
sapply(diabetes_factor, class)
```
Next provide a quick summary of each variable
```{r}
summary(diabetes_factor)
```
Plotting all variables with an instruction is a bit tricky. May be easier to plot separately numerical and categorical variables.

```{r}
library(ggplot2)
library(tidyr)
library(purrr)

diabetes %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```

Proceed similarly with categorical variables

```{r}
diabetes %>%
  keep(is.character) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_bar()
```

You may notice -or not- that the dataset has some outlier values.

Before removing them consider estimating the mean nvalue of SBP and DBP with distinct estimators

```{r}
with(diabetes_factor, {
    print("DBP")
    show(summary(dbp))
    print("SBP")
    show(summary(sbp))
  }
  )
```
What is prefereable to estimate the mean SBP or DBP?



## Example 2. Computing Confidence Intervals with R (2)

```{r}
t.test(osteo100[["bua"]])
```
---

## Example 2 . Computing Confidence Intervals with R (3)

```{r}
cntMenop <- table(osteo100[["menop"]])["SI"]
ssize <- length(osteo100[["menop"]])
prop.test (x=cntMenop, n=ssize)
```

# Exercise 2.1 Computing Confidence intervals

- Read the file "osteoporosis.csv" into a dataset and call it "osteoporosis"

- Compute confidence intervals for the BUA mean and for the percentage of menopausic women with __all the individuals in the dataset__.

- Compare these confidence intervals with those that you obtained in example 2. How do they differ?



# Exercise 2.2 Computing Confidence intervals

- Read the diabetes dataset. Convert characters into factors before continuing.

- Provide a confidence interval for:
    - a proportion of at least one categorical variable and 
    - the mean value of at least one numerical variable.

- How would you find alternative approaches to compute these confidence intervals?

- Why would you want to do such a thing?


# Example 3. Sample size calculation

- Using the osteoporosis dataset, assume that the standard deviation is a good aproximation to $\sigma$.

- Find the sample size needed to achieve a margin of error equal to 5 with a $95\%$ confidence interval.

# Exercise 3. Sample size calculation

- Write a function to compute the sample size for proportions in the worst case (p=q=0.5) or assuming $p$ is known.

- Using a $50\%$ planned proportion estimate, find the sample size needed to achieve $5%$ margin of error for a survey at $95%$ confidence level.

- How would this result change if we are told that a pilot study suggests that $p=10\%$?




