---
title: "Statistics with R. Intro to hypothesis testing"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Hypothesis testing examples

Researchers studying osteoporosis have some previous belief about the characteristics of a "typical" population:

Examples of such beliefs are:

- The average ''bua'' value in our population is 70.
- The ''bua'' mean value in menopausic and non-menopausic women is not the same.

We will use the osteoporosis dataset to check these beliefs. To make it more interesting/realistic we will assume we have small sample sizes and __everyone will have a different sample__

```{r}
osteoporosis <- read.delim2("datasets/osteoporosis.csv", stringsAsFactors=TRUE)
```

We will select a subset of this dataset which is different for everyone.

```{r}
library(dplyr)
semilla<- 18101998
set.seed(semilla)
numMuestras <- 25
muestras <- sample( osteoporosis$registro, numMuestras)
osteoData <- dplyr::filter(osteoporosis, registro  %in%  muestras)
```

We have used the `filter` function from the ` dplyr` package to select rows.
If we wanted to keep only a few columns we would have used the ` select` function.

```{r}
osteoData2 <- dplyr::select(osteoData, c("edad", "imc", "bua", "menop"))
```

Instead of creating several datasets we can use the "pipe" operator.

```{r}
osteoData <- osteoporosis %>%  
  filter(registro  %in%  muestras) %>%
  select(c("edad", "imc", "bua", "menop"))
```

## Answering the questions:

- _The average ''bua'' value in our population is 70?_

```{r}
m <- mean(osteoData$bua)
print(m)
```
Use a confidence interval to decide about your hypothesis:

```{r}
t.test(osteoData$bua)
```
- _The ''bua'' mean value in menopausic and non-menopausic women is not the same._

We could filter twice our subset and compute the means separately.
An easier way is to use the `group_by` and the `summarize` operators:

```{r}
osteoData %>% 
  group_by(menop) %>% 
  summarize(m=mean(bua), s=sd(bua), l=length(bua))

```

## Computing the critical value for a test

Assuming we do a test for the mean, with an unknown standadr deviation the critical value for a one-sided hypothesis is: 
$$
m_c=\overline X + t_{\alpha, n}\times s/\sqrt{n}
$$
In our example:

```{r}
mc <- mean(osteoData$bua) + qt(0.95, numMuestras-1) * sd(osteoData$bua)/sqrt(numMuestras)
print(mc)
```

If our sample mean is smaller than we ```r mc```` then the null is rejected

If the population variance is known the formula does not depend on the sample size
$$
m_c=\overline X + z_{\alpha}\times \sigma/\sqrt{n}
$$
In this case the critical value is:
```{r}
alfa=0.05
sigma <- sd(osteoporosis$bua)
mc <- mean(osteoData$bua) + qnorm(1-alfa) * sigma/sqrt(numMuestras)
print(mc)
```

## Computing the p-values

```{r}
pval<- pnorm(q=74, mean=70, sd=17, lower.tail=FALSE)
print(pval)

```

```{r}
pval<- pnorm(q=85, mean=70, sd=17, lower.tail=FALSE)
print(pval)

```

## Testing  hypothesis abot BUA using built-in tests

```{r}
t.test(osteoData$bua, mu=70, alternative="less")
```


```{r}
t.test(bua~menop, mu=70, data=osteoData)
```



## Exercises

### Practice filtering

- Read the `diabetes`` dataset and create a subset made by non smokers who are older than 50 years old.
- Take the previous dataset and keep only three variables: "edad", "sbp" and "dbp"
- Did you use the "pipe" operator. If not, repeat the exercises in a single step using it.

# Power calculations tutorial

This tutorial shows how to perform power and sample size calculations
in R for the case where the outcome variable is either continuous or
binary.


# Binary outcome

Binary outcome means that every subject has either (1= event) or (0=
no event). The event probability is the probability that a subject has
the event. The aim is to show that two groups differ with respect to
the event probability.


## Power for given sample size

Aim: to compute the power of a study to show a difference between
group 1 (n=28) in which the event probability is 30% and group 2
(n=28) in which the event probability is 55%.

```{r   }
power.prop.test(n=28,p1=0.3,p2=0.55) 
```

Possible conclusion sentence:

*The power of a study which includes 28 subjects in each of two experimental groups to see a difference between the event probabilities is 48% under the assumption that the event probabilities are 30% in group 1 and 55% in group 2.*


## Sample size for given power

Aim: To compute the sample size of a study to show a difference
between group 1 (n=28) in which the event probability is 30% and group
2 (n=28) in which the event probability is 55% with a power of 80%.

```{r   }
power.prop.test(power=0.8,p1=0.3,p2=0.55) 
```

Possible conclusion sentence:

*In order to achieve a power of 80% under the assumed event probabilities the study should include at least 61 subjects in each of the experimental groups.*


# Continuous outcome

A continuous outcome variable is a variable that can take on any value
between its minimum and its maximum value, such as weight or BMI or
the blood hemoglobin concentration. For the power and sample size
calculations illustrated here the distribution of the outcome variable
is characterized by its mean and standard deviation.


## Power for given sample size

Aim: to compute the power of a study which aims to show a difference
in means between group 1 (n=6) and group 2 (n=6) assuming that the
magnitude of the difference is 0.3 units and the standard deviation is
0.28 units.

```{r   }
power.t.test(n=6,delta=0.3,sd=0.28,type="two.sample") 
```

Possible conclusion sentence: 

*The power of the study is 39% to detect a difference in means of 0.3 units.*


## Sample size for a given power

Aim: to compute the sample size needed to achieve a power of 90% in a
study which aims to show a difference in means between two independent
groups assuming that the magnitude of the difference is 0.3 units and
the standard deviation is 0.28 units.

```{r   }
power.t.test(power=0.9,delta=0.3,sd=0.28,type="two.sample") 
```

Possible conclusion sentence: 

*To reach a power of 90% the study should include at least 20 subjects in each group to detect a difference in means of 0.3 units.*


# Plotting power against sample size

The functions `power.prop.test` and `power.t.test` can compute the
power for a sequence of sample sizes simultaneously.

The following code shows how to make a graph which shows the increase
of the power for a given sequence of sample sizes

```{r   }
samplesizes <- seq(from=10,to=200,by=10)
power.samplesizes <- power.prop.test(n=samplesizes,p1=0.3,p2=0.55)$power
plot(samplesizes,
     power.samplesizes,
     xlim=c(0,200),
     xlab="Sample size",
     ylab="Expected power",
     ylim=c(0,1),
     type="b",
     col="darkorange",
     lwd=5,axes=FALSE)
axis(1,at=c(0,50,100,150,200))
axis(2,at=c(0,0.25,0.5,0.75,1),labels=paste(c(0,25,50,75,100),"%")) 
```

Often the greatest concern is the magnitude of the expected difference
between the groups, even if based on historical data or a pilot
study. Therefore a useful plot shows how the sample size for fixed
power (or power for fixed sample size) varies as a function of the
difference. However, the functions `power.prop.test` and
`power.t.test` can unfortunately not deal with a sequence of
differences. Therefore we use a loop (`sapply`) in the following
example. To show the effect of the hypothesized standard deviation we
also show the required sample size when the standard deviation is
only 0.2 instead of 0.4.

```{r   }
differences <- seq(from=0.1,to=2,by=0.1)
samplesize.sd04 <- sapply(differences,
                          function(d){power.t.test(power=0.9,
                                                   delta=d,
                                                   sd=0.4,
                                                   type="two.sample")$n})
samplesize.sd03 <- sapply(differences,
                          function(d){power.t.test(power=0.9,
                                                   delta=d,
                                                   sd=0.3,
                                                   type="two.sample")$n})
plot(differences,
     samplesize.sd04,
     xlim=c(0,2),
     xlab="Expected difference between groups",
     ylab="Required sample size",
     ylim=c(0,350),
     type="b",
     col="darkblue",
     lwd=5,axes=FALSE)
lines(differences, samplesize.sd03,col="turquoise",lwd=5,type="b")
axis(1,at=c(0,0.2,0.5,1,1.5,2))
axis(2,at=c(350,100,50,10,0))
legend(x="topright",lwd=5,bty="n",legend=c("SD=0.4","SD=0.3"),col=c("darkblue","turquoise")) 
```







