---
title: "Statistics with R. Intro to hypothesis testing"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

# Case Study: The `osteoporosis` dataset

Researchers studying osteoporosis have some previous belief about the characteristics of a "typical" population:

Examples of such beliefs are:

- The average ''bua'' value in our population is 70.
- The ''bua'' mean value in menopausic and non-menopausic women is not the same.

We will use the osteoporosis dataset to check these beliefs. To make it more interestin/realistic we will assume we have small sample sizes and __everyone will have a different sample__

```{r}
osteoporosis <- read.delim2("/datasets/osteoporosis.csv", stringsAsFactors=TRUE)
```

We will select a subset of this dataset which is different for everyone.

```{r}
library(dplyr)
semilla<- 18101998
set.seed(semilla)
numMuestras <- 25
muestras <- sample( osteoporosis$registro, numMuestras)
osteoData <- dplyr::filter(osteoporosis, registro  %in%  muestras)
```

We have used the `filter` function from the ` dplyr` package to select rows.
If we wanted to keep only a few columns we would have used the ` select` function.

```{r}
osteoData2 <- dplyr::select(osteoData, c("edad", "imc", "bua", "menop"))
```

Instead of creating several datasets we can use the "pipe" operator.

```{r}
osteoData <- osteoporosis %>%  
  filter(registro  %in%  muestras) %>%
  select(c("edad", "imc", "bua", "menop"))
```

## Answering the questions:

- _The average ''bua'' value in our population is 70?_

```{r}
m <- mean(osteoData$bua)
print(m)
```
Use a confidence interval to decide about your hypothesis:

```{r}
t.test(osteoData$bua)
```
- _The ''bua'' mean value in menopausic and non-menopausic women is not the same._

We could filter twice our subset and compute the means separately.
An easier way is to use the `group_by` and the `summarize` operators:

```{r}
osteoData %>% 
  group_by(menop) %>% 
  summarize(m=mean(bua), s=sd(bua), l=length(bua))

```

## Computing the critical value for a test

Assuming we do a test for the mean, with an unknown standadr deviation the critical value for a one-sided hypothesis is: 
$$
m_c=\overline X + t_{\alpha, n}\times s/\sqrt{n}
$$
In our example:

```{r}
mc <- mean(osteoData$bua) + qt(0.95, numMuestras-1) * sd(osteoData$bua)/sqrt(numMuestras)
print(mc)
```

If our sample mean is smaller than we ```r mc```` then the null is rejected

If the population variance is known the formula does not depend on the sample size
$$
m_c=\overline X + z_{\alpha}\times \sigma/\sqrt{n}
$$
In this case the critical value is:
```{r}
alfa=0.05
sigma <- sd(osteoporosis$bua)
mc <- mean(osteoData$bua) + qnorm(1-alfa) * sigma/sqrt(numMuestras)
print(mc)
```

## Computing the p-values

```{r}
pval<- pnorm(q=74, mean=70, sd=17, lower.tail=FALSE)
print(pval)

```

```{r}
pval<- pnorm(q=85, mean=70, sd=17, lower.tail=FALSE)
print(pval)

```

## Testing  hypothesis abot BUA using built-in tests

```{r}
t.test(osteoData$bua, mu=70, alternative="less")
```


```{r}
t.test(bua~menop, mu=70, data=osteoData)
```







# Exercises

## Practice filtering

- Read the `diabetes`` dataset and create a subset made by non smokers who are older than 50 years old.
- Take the previous dataset and keep only three variables: "edad", "sbp" and "dbp"
- Did you use the "pipe" operator. If not, repeat the exercises in a single step using it.






