---
title: "EDA with R"
output: html_notebook
---

## The Data Science Approach in R

# Getting started

## Getting started (I)

1.  Load the dataset \emph{diabetes}:

```{r, size="footnotesize", message=FALSE}
library(readxl)
diab <- read_excel("diabetes.xls")
```

2.  Check if we have loaded it correctly:

```{r, size="tiny"}
diab[1:4, 1:8]
```


## Getting started (II): functions to check a dataframe:
-   Content

    - `head`(): shows the first few rows 
    - `tail`(): shows the last few rows

-   Size

    - `dim`(): returns the number of rows and the number of columns 
    - `nrow`(): returns the number of rows 
    - `ncol`(): returns the number of columns

-   Summary

    - `colnames`() or names(): returns the column names 
    - `glimpse`(): returns a glimpse of your data: structure, class, length and content of each column


## Getting started (III)

```{r size="tiny"}
head(diab)
```

## Getting started (IV)

```{r}
dim(diab)
nrow(diab)
colnames(diab)
```

## Getting started (IV)

```{r size="tiny"}
library(tidyverse)
glimpse(diab)
```

## Variables and data types

- Data managed in R ...

  - is stored as *variables*

- Variables can be of distinct types

  - Numerical

    -  numeric (13.7)
    -  int (3)

  - Character

     - "R is cute"

  - Factors are used to represent _categorical data_

    - A,B,C,D
    - WT, Mut

  - Logical

    -  TRUE/FALSE

## Exercise I

- Load the osteoporosis dataset

- Proceed similarly as to what we have done above and obtain information on

  - How many variables and observations
  - How are them

## More about factors

-   Each data type is what it seems to be, but factors require more explanation.

-   Factors are intended to describe categories such as "sex", "blood group", but also "risk" or "stage".

-   Factors are useful to describe groups without having to use numeric codes.

-   Factors may be created while reading the file or later using the `factor` and `as.factor` commands.

## Create factor while reading

- Factors may be created automatically from string data when

  - Files are read using the `read.delim`, `read.csv`  or `read.table` functions of R base
  - Option `stringAsFactors' is set to TRUE

- Reading files from Excel (`read_excel`) or using `tidyverse` functions such as `read_csv` does not allow to  create factors automatically.

## Read a file and create factors automatically

- Import the diabetes dataset from the `diabetes.csv` file using the Rstudio dialog.

```{r}
diabetes <- read.csv("datasets/diabetes.csv", stringsAsFactors=TRUE)
glimpse(diabetes)
```

## Read a file without creating factors automatically

- Re-read the file 

  - from excel (`diabetes.xls`)
  - or without setting the "stringsAsFactors" to TRUE

```{r}
diab <- read.csv("datasets/diabetes.csv", stringsAsFactors=FALSE)
glimpse(diab)
```

## Creating factors directly

- Use `factor` or `as.factor`

```{r}
diabetes <- read.csv("datasets/diabetes.csv", stringsAsFactors=FALSE)
class(diabetes$mort)
diabetes$mort <- as.factor(diabetes$mort)
class(diabetes$mort)
levels(diabetes$mort)
```

**Warning!** by default alphabetic order is used when creating factor levels.

```{r}
vitalStatus <- factor(diabetes$mort, levels=c("Vivo", "Muerto"))
class(vitalStatus)
levels(vitalStatus)
```

## Change the levels of a factor

- When humans fill the database... many errors can happen :(

  - An answer like "YES", could be entered like: 
    
          "YES", "yes", "Yes", "Yeah "

- All this possible answers **will be different levels for the same variable**

- This may be solved using `recode_factor`:

```{r}
diab$mort <- recode_factor(diab$mort, "Muerto" = "muerto")
levels(diab$mort)
```

## Changing *characters (chr)* to *factors (Factor)*

An alternative way to turn characters into factors is the  `mutate_if` function:

```{r size="tiny"}
library(dplyr)
diab <- diabetes %>% mutate_if(is.character, as.factor)
glimpse(diab)
```

## Creating factors when categorizing numerical variables

- `diab$edat` is a numerical variable but we may want to categorize it that is creat a new factor variable whose categories are defined by nuemric intervals.

```{r}
grupsEdat <- cut(diab$edat, breaks=c(0,30,50,75,Inf),
                include.lowest=TRUE,
                labels=c("<30","30-50","50-75",">75"))
library(tibble)
diab <- diab %>% add_column (grupsEdat, .after="edat")
glimpse(diab[,2:5])
```

# Descriptive Statistics: Numerical summaries

## Numerical Summaries (I)

- There are many functions to provide numerical summaries

```{r size="tiny"}
#Mean, median and rang
mean(edat)
mean(diab$edat)
median(diab$edat)
sd(diab$edat)
var(diab$edat)
range(diab$edat)
attach(diab)
mean(edat)
detach(diab)
```

## Numerical Summaries (II)

A general summary of all variables is provided by distinct functions

```{r size="tiny"}
summary(diab)
summary(diab[, 2:11])
```

## Numerical SUmmaries (III)

If categorical variables are not adequately represented summaries will be less informative.

Remember that we read created `diabetes` without turning variables into factors.

```{r}
diabetes <- read.csv("datasets/diabetes.csv", stringsAsFactors=FALSE)
summary(diabetes)
```

## Improving the `summary` function

- There are many packages to do descriptive statistics.

- See [Dabbling with data](https://dabblingwithdata.wordpress.com/2018/01/02/my-favourite-r-package-for-summarising-data/)

- Give a try, for instance to the `skimr` or `summarytools` packages.

## More complete descriptions (I)

```{r}
library(summarytools)
summarytools::dfSummary(diabetes)
```

## Grouped summaries

If we want to group the descriptive summaries by other variables we can use `group_by` function:

```{r size="tiny"}
diab %>% 
  group_by(tabac, ecg) %>%
  summarize(mean(edat))
```

## Handling missing data

- What happens if we have missing data in our dataset?

- The file `diabetes_mod.xls` contains some missings

```{r size="tiny"}
diabetes_mod <- read_excel("datasets/diabetes_mod.xls")
diab <- diabetes_mod %>% mutate_if(is.character, as.factor)
mean(diab$sbp)
```

**NA** indicates *missing data* in the variable

Let's look the `sbp` variable:

```{r size="tiny"}
diab$sbp
```

## How to work with *missing data*:

```{r size="tiny"}
?mean
mean(diab$sbp, na.rm = TRUE)
is.na(diab$sbp)
```

## How to work with *missing data*:

- Count missings

```{r}
sum(is.na(diab$sbp))
sum(is.na(diab$dbp))
```

- Remove all rows with missing
```{r}
diab_noNAS <-na.omit(diab)  
dim(diab)
dim(diab_noNAS)
```

See also: [Remove Rows with NA in R Data Frame](https://statisticsglobe.com/r-remove-data-frame-rows-with-some-or-all-na)

## EXERCISE II

With the `diab` dataset

-   Show only the rows from 35 to 98 and columns 5, 7, and from 9 to 11

-   Change the level of the variable *tabac*, from **No Fumador** to **No\_Fumador**

-   Display the unique values for the variable *bmi*. Count how many exist.

-   Display the mean of *edatdiag*, grouped by *ecg*


## Exercise III

  - Read the `osteoporosis.csv` data set  into R so that character variables are automatically converted into factors.
  
  - Make a summary of the dataset using standard and non-standard summary functions.
  
  - Display the range of `bua`, grouped by `clasific`
  
## Tuning plots with graphic parameters

The aspect of plots can be improved using "graphical parameters"

Some parameters are passed as arguments to the plotting function

Others are set globally.

Setting graphical parameters has become so sophisticated that a new approach to graphics -in principle more intuitive- has emerged and come to change the way that graphics are created.

- The "traditional" approach relies on the `plot` function available in Rbase
- The "modern" approch is based on the `ggplots2` package and its "language" of graphics.

## An improved histogram 

```{r}
hist(diab$edat, main="Age at diagnostic of diabetes",
     probability = TRUE, xlab="Age", ylab="Percentage", 
     col="lightblue", breaks = 10)
```


## Boxplot. A unidimensional histogram

```{r}
boxplot (diab$edat, main ="Age")
```

## Boxplot: Decomposing plots by groups

```{r}
boxplot(edat~tabac, data=diab, main="Age of diabetes onset according smoking status")
```


## Plots for categorical variables

- Some simple principles

  + Use pie charts only with categorical variables in nominal scale
  + Use barplots for any categorical variable
  + Never use 3D-plots

```{r}
t <- table(diab$mort)
pie(t)
```

## Barplots

- Similar to pie charts but, implicitly, suggest ordering

```{r}
barplot(t)
```

## Scatterplots

- For two variables, simply use `plot`

```{r}
plot(edat~bmi, data=diab)
```

## Exercise IV

- With the variables in the osteoporosis dataset

- Try to represent the different variables using _the most appropriate plot_ for each of them.