---
title: "EDA with R"
output: html_notebook
---

# Outline: Exploratory Analysis (II)

- Introduction: From 1 to 2 or more diomensions

- Exploring relation between variables

- Introducing Correlation


# From one to more dimensions

- A natural way to start looking at data is, as we have done, on a "one variable-at-a-time" approach.

  - Reasonable way to start understanding the data, but
    
  - Fails at the possibility of detecting relations between variables, which can be the basis to understand what we had seen before.

## Why are bi/multivariate approaches better?

- Exploring two ore more variables simultaneosuly can provide more information

  - Does  there exist a relation between certain variables?
    
  - How important is this relation?
    
  - Which is the direction of the relation?
  

# Elegant Graphics for data analysis

```{r}
#install the package
#install.packages(ggplot2)
#load the package
library(ggplot2)
```


```{r}
#see the data
head(mpg)
colnames(mpg)
```


```{r}
#do the basic plot
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```


```{r}
#change to lines
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_line()
```


```{r}
#Improving the graphic (color by manufacturer)
ggplot(mpg, aes(x = displ, y = hwy, color = manufacturer)) + 
  geom_point()
```


```{r}
#Improving the graphic II (add shape by drv)
ggplot(mpg, aes(x = displ, y = hwy, color = manufacturer, shape =  drv)) + 
  geom_point()
```

# Exercise 1

- Load the diabetes dataset

- Draw a scatterplot showing the relation between `bmi` and `sbp`

# Bivariate Analysis

## Qualitative versus qualitative

```{r}
#load the data 
osteoporosis <- read.csv2("datasets/osteo100.txt", sep = "\t", header = TRUE, dec = ".", stringsAsFactors = TRUE)

#see the data is correctly loaded
head(osteoporosis)
```


```{r}
#contingency table
tabEdatClassific <- table(osteoporosis$grupedad, osteoporosis$clasific)
tabEdatClassific
```


```{r}
#contingency table in %
prop.table(tabEdatClassific)
```

```{r}
barplot(tabEdatClassific, beside = TRUE)
```
```{r}
barplot(t(tabEdatClassific), beside = TRUE)
```



```{r}
#plot the data: stacket barplot
ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific))
```


```{r}
#plot the data II: bars side by side
ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific), position = "dodge")
```


```{r}
#Change colors, legend position, labels and finally save it!
p <- ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific), position = "dodge")

p + scale_fill_manual(values=c("#8618b1", "blanchedalmond", "red"))

p + theme(legend.position = "bottom")

p + labs(x = "Age group", y = "Women", title = "Osteo disease classified by age group")

pdf("clasific_grupedad.pdf")
  p + labs(x = "Age group", y = "Women", title = "Osteo disease classified by age group")
dev.off()
```

### Another way to introduce the data

```{r}
tab <- matrix(data = c(120, 60, 50, 70), nrow = 2, ncol = 2, byrow = TRUE)
tab
```


```{r}
#change colnames and rownames
colnames(tab) <- c("Smokers", "Nonsmokers")
rownames(tab) <- c("Men", "Women")
tab
```


```{r}
#Look in %
prop.table(tab)
```

## Qualitative versus quantitative

Tabulation by groups is a bit weird with Rbase

```{r}
#Table of statistics
with(osteoporosis, tapply(bua, list(grupedad), mean, na.rm=TRUE))
```

But it is more natural using a tidyverse approach

```{r size="tiny"}
library(dplyr)
osteoporosis %>% 
  group_by(grupedad) %>%
  summarize(mean(edad))
```


```{r}
#Plot the data
bp <- ggplot(osteoporosis, aes(x = grupedad, y = bua)) + 
  geom_boxplot(fill = '#A4A4A4', color = "darkred")
bp

# Box plot with points
# 0.2 : degree of jitter in x direction
bp + geom_jitter(shape = 16, position = position_jitter(0.2)) +
   labs(x = "Age Group", y = "Women", title = "Osteo disease classified by age group")
```

## Quantitative versus quantitative

```{r}
# Basic scatter plot
ggplot(osteoporosis, aes(x = peso, y = imc)) + 
  geom_point()

# Change the point size, and shape
ggplot(osteoporosis, aes(x = peso, y = imc)) +
  geom_point(size = 1, shape = 1)

# Color the points depending of another variable
ggplot(osteoporosis, aes(x = peso, y = imc, color = clasific, shape = clasific)) +
  geom_point()

#not always the correlation is good
ggplot(osteoporosis, aes(x = edad, y = imc)) + 
  geom_point()
```


```{r}
#correlation matrix
pairs(osteoporosis[, c("edad", "peso",  "talla", "imc", "bua", "menarqui")])
```


```{r}
#with ggplots
#install.packages(GGally)
library(GGally)
                 
ggpairs(osteoporosis, columns = c("edad", "peso",  "talla", "imc", "bua", "menarqui"), ggplot2::aes(colour = clasific))

```

# Correlation

```{r}
#Pearson correlation
cor(osteoporosis$bua, osteoporosis$edad, method = "pearson")
```


```{r}
#the plot

ggplot(osteoporosis, aes(x = edad, y = bua)) + 
  geom_point()

#Spearman correlation
cor(osteoporosis$bua, osteoporosis$edad, method = "spearman")
```


  
