---
title: "Exploratory Analysis with R (II)"
author: "Alex Sanchez, Miriam Mota, Ricardo Gonzalo and  \n
        Santiago Perez-Hoyos"
date: "Statistics and Bioinformatics Unit. Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: "Madrid"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    slide_level: 2
footer: "Statistics with R"
editor_options: 
  chunk_output_type: console
---

```{r  include=FALSE}
require(knitr)
opts_chunk$set(
concordance=FALSE, echo=TRUE, cache=FALSE, warning=FALSE, error=FALSE, message=TRUE, fig.height=4, size = "footnotesize")

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

```{r setLicense, child = 'license.Rmd', eval =F}
```

## Outline: Exploratory Analysis (II)

- Introduction: From 1 to 2 or more diomensions

- Exploring relation between variables

- Introducing Correlation


## From one to more dimensions

- A natural way to start looking at data is, as we have done, on a "one variable-at-a-time" approach.

  - Reasonable way to start understanding the data, but
    
  - Fails at the possibility of detecting relations between variables, which can be the basis to understand what we had seen before.

## Why are bi/multivariate approaches better?

- Exploring two ore more variables simultaneosuly can provide more information

  - Does  there exist a relation between certain variables?
    
  - How important is this relation?
    
  - Which is the direction of the relation?
  

  
