---
title: "Basic Statistics for Biomedical Research"
author: Santiago Perez-Hoyos, Miriam Mota, Marcos Esteve
date: "Statistics and Bioinformatics Unit.\n Vall d'Hebron Institut de Recerca"
output:
  beamer_presentation:
    theme: Copenhagen
    colortheme: dolphin
    fonttheme: structurebold
  slide_level: 2
  ioslides_presentation: default
  slidy_presentation: default
footer: Basic Statistics for Biomedical Research
editor_options:
  markdown:
    wrap: 72
---

```{r setLicense, child = 'license.Rmd'}
```

# Introduction

## Outline

-   Introduction

    -   Who are we ("we"=teachers & students)
    -   Why are we here (Why learn R?)

-   How will we proceed: Methodology

-   HW Data Science approach to using R

-   References & Resources

## Who are we: The Statistics and Bioinformatics Unit

[https://uebvhir.github.io/WUEB/docs/](https://uebvhir.github.io/WUEB/docs/)

```{=tex}
\begin{figure}
\includegraphics[width=0.95\linewidth]{images/sessio0_1.png}
\end{figure}
```
## What we do: Teachers

```{=tex}
\begin{figure}
\includegraphics[width=0.95\linewidth]{images/sessio0_2.png}
\end{figure}
```
<!-- ## Who are we (3): The GRBio Research group -->

<!-- \begin{figure} -->

<!-- \includegraphics[width=0.85\linewidth]{images/GRBio.jpg} -->

<!-- \end{figure} -->

## Why this course (1)

```{=tex}
\begin{figure}
\includegraphics[width=0.5\linewidth]{images/drowningindata.jpg}
\end{figure}
```
"We are drowning in information but starving for knowledge"

## Why this course (2)

-   **(Biomedical) research**, as well as many other human activities
    (social networks, sports, COVID ...) generate **huge quantities** of
    -often complex- **data.**

    -   Although sometimes we will also have small datasets

-   We believe that data leads to information that leads to knowledge,
    but we need to be able to extract one from the other.

-   This can be attempted in many ways, artificial intelligence, machine
    learning, data science or something which is common to all of them:
    **Plain Statistics**!!!

## What are our goals (1)

-   The main goal of this course is to introduce a variety of
    statistical methods and tools, which is good enough to:

    -   Help you analyze your own data when it makes sense
    -   Suggests you when the analysis is complex enough to contact an
        expert statistician (such us those in the UEB)
    -   Help you to distiguish one from the other

<p>

</p>

-   A secondary, but not least important objective: Show how to do it
    using R


```{r setLicense, child = 'license.Rmd'}
```

# Introduction

## Outline

\scriptsize

| Sessió | Data     | Modalitat   | Professor | Títol                                                               |
|--------|----------|-------------|-----------|---------------------------------------------------------------------|
| 1      | 20/05  | En línia    | Santi     | Introducció a R i RStudio i Tidyverse                                 |
| 2      | 21/05  | En línia    | Miriam    | Estadística descriptiva I: Resums numèrics, taules i gràfics.         |
| 3      | 22/05  | En línia    | Miriam    | Estadística descriptiva II: Gràfics i taules bivariants. ggplot2      |
| 4      | 27/05  | En línia    | Miriam    | Més manegament de dades amb R. Automatització amb scripts             |
| 5      | 29/05  | Presencial  | Santi     | Introducció a la inferència estadística. Intervals de confiança       |
| 6      | 3/06   | En línia    | Santi     | Proves d’hipòtesis I: Conceptes bàsics                                |
| 7      | 4/06   | En línia    | Santi     | Proves d’hipòtesis II: Variables quantitatives                        |
| 8      | 5/06   | En línia    | Miriam    | Proves d’hipòtesis III: Variables qualitatives                        |
| 9      | 10/06  | En línia    | Miriam    | Tests diagnòstics: Sensibilitat, especificitat i corbes ROC           |
| 10     | 12/06  | Presencial  | Santi     | Exercici resum d’anàlisis de dades reals                              |

**Horari:** 13:00h a 15:00h  
**Sessions presencials:** *Sala de Juntes*  
**Sessions en línia:** *via plataforma virtual*

\normalsize

## Course contents

\footnotesize

-   Sessió 1. Introducció a R i RStudio i Tidyverse.

-   Sessió 2. Estadística descriptiva I: Resums numèrics, taules i
    gràfics. Reproduïbilitat amb R

-   Sessió 3. Estadística descriptiva II: Gràfics i taules bivariants.
Visualització de dades amb ggplot2.

-   Sessió 4. Més manegament de dades amb R. Automatització de tasques
    amb scripts..

-   **Sessió 5. Sessió presencial de pràctiques en R i resolució de
    dubtes**

-   Sessió 6. Introducció a la inferència. Intervals de Confiança.

-   Sessió 7. Proves d'hipòtesis I: Conceptes bàsics

-   Sessió 8. Proves d'hipòtesis II: Variables quantitatives.

-   Sessió 9. Proves d'hipòtesis III: Taules de contingència, Khi²,
    Tests diagnòstics: Sensibilitat, especificitat i corbes ROC.

-   **Sessió 10. Sessió presencial de resolució d’exercici d’anàlisis de
    dades reals.**

-   If there is anything else you would like to learn, let us know and
    we'll try not to let it out.
    
\normalsize

## Why doing statistics with R

-   R has become a "de facto" standard for statistical analysis
-   R is free to use!!
-   Practically all existing statistical methods available
-   Powerful graphics that can be used interactively to explore data
-   Possibility of scripting analysis $\longrightarrow$
    **Reproducibility**
-   Enormous and collaborative Community

## How we will work

1.  Our session will have the following structure (all but the first)

    -   1st part: We introduce a few new theoretical ideas

    -   2nd part: We wil show how to do it in R

        3rd part: Practice exercises and start working on the case study
        suggested or if you want in your data.

2.  Mastering R requires as many other disciplines

    -   Time

    -   Study, and

    -   Practice.

So take it easy. You can go back to R once the course finish

Class atendance and participation (class atendance should be at least
80%)

## The materials

You can find the presentations, scripts , exercises in the teaching
section of our web ueb.vhir.org

```{=tex}
\begin{figure}
\includegraphics[width=0.95\linewidth]{images/sessio0_4.png}
\end{figure}
```
## - Resources and references

There is a huge variety of resources to learn R, books, tutorials, free
online courses, etc.

-   This course is based on the book [Data Science for
    R](http://r4ds.had.co.nz/).

-   Other interesting books + [Using R and RStudio for Data Management,
    Statistical Analysis, and Graphics, 2nd
    edition](https://www.amazon.com/Management-Statistical%20-Analysis-Graphics-2015-04-27/dp/B017POAKSK/ref=sr_1_5?ie=UTF8&qid=1539194380&sr=8-5&keywords=using+r+and+rstudio+for+data+management)

-   Online courses + [Coursera's Data Science
    Specialization](https://github.com/DataScienceSpecialization/courses)

-   [A list of R tutorials and
    courses](https://hackr.io/tutorials/learn-r)

## 

```{=tex}
\begin{figure}
\includegraphics[width=1\linewidth]{images/sessio0_5.png}
\end{figure}
```
