---
title: "StatisticsWithR-2-EDA-Exercises.Rmd"
author: "UEB"
# date: "4/6/2021"
output: html_document
---
## Exercise I

-   Load the `osteoporosis` dataset

```{r}

```


-   Proceed similarly as to what we have done above and obtain information on

    -  How many variables and observations
    -  How are them (What are their types)
    
```{r}

```
    
  
## EXERCISE II

1.  With the `diab` dataset

-   Show only the rows from 35 to 98 and columns 5, 7, and from 9 to 11

-   Change the level of the variable *tabac*, from **No Fumador** to **No\_Fumador**

-   Display the unique values for the variable *bmi*. Count how many exist.

-   Display the mean of *edatdiag*, grouped by *ecg*


```{r}

```



## Exercise III

- Using the *osteoporosis.csv* dataset

 - Read the `osteoporosis.csv` data set  into R so that character variables are automatically converted into factors.
 - Load the dataset and check if it is correctly loaded
 - Make a numerical summary of the dataset.
 - Calculate the mean and standard deviation of imc grouped by clasific
 - Plot the distribution of edat

```{r}

```

  
## Exercise IV

- With the variables in the osteoporosis dataset

```{r}

```


- Try to represent the different variables using _the most appropriate plot_ for each of them.

```{r}

```

