---
title: "Statistics Course with R - Day 3"
author: "UEB"
date: "27/04/2021"
output: pdf_document
toc: true
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Elegant Graphics for data analysis

```{r}
#install the package
#install.packages(ggplot2)
#load the package
library(ggplot2)
```

```{r}
#see the data (we'll take data from package)
head(mpg)
colnames(mpg)
str(mpg)
summary(mpg)
```

```{r}
#do the basic plot
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```

```{r}
#it can be assigned to an object too
p <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
p
```

```{r}
#change geom to lines
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_line()
```

```{r}
#add layers
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_line()
```

```{r}
#add layers (title)
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_line() +
  labs(title="Plot of mpg data", x="displ(l)", y="hwy(miles/gallon)")
```

```{r}
#formatting labels
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_line() +
  labs(title="Plot of mpg data", x="displ(l)", y="hwy(miles/gallon)") +
  theme(plot.title=element_text(face="bold", hjust=0.5))
```

```{r}
#Playing with aes (color by class)
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + 
  geom_point()
```

```{r}
#Playing with aes (add shape by drv)
ggplot(mpg, aes(x = displ, y = hwy, color = class, shape =  drv)) + 
  geom_point()
```

```{r}
#Facets
ggplot(mpg, aes(x = displ, y = hwy, color = class)) + 
  geom_point() +
  facet_grid(. ~ drv)
```

# Descriptive statistics for univariate data

## Quantitative variables

* Numeric summaries

```{r}
mean(mpg$displ)
median(mpg$displ)
sd(mpg$displ)
#using summary function
summary(mpg$displ)
```

* Graphic summaries

```{r}
hist(mpg$displ)
```

```{r}
boxplot(mpg$displ)
```

With ggplot

```{r}
ggplot(mpg, aes(displ)) +
  geom_histogram()
ggplot(mpg, aes(displ)) +
  geom_boxplot()
ggplot(mpg, aes(displ)) +
  geom_boxplot() + 
  coord_flip()
```

## Qualitative variables

* Numeric summaries

```{r}
#absolute frequencies
table(mpg$class)

#relative frequencies
prop.table(table(mpg$class))
```

```{r}
# install.packages("gmodels")
library(gmodels)
CrossTable(mpg$class)
```

* Graphic summaries

```{r}
barplot(table(mpg$class))
barplot(prop.table(table(mpg$class)))
```

With ggplot2

```{r}
ggplot(mpg, aes(class)) +
  geom_bar()
```


```{r}
#for relative frequency
ggplot(mpg, aes(class)) +
  geom_bar(aes(y=..prop.., group=1))

ggplot(mpg, aes(class)) +
  geom_bar(aes(y=..prop.., group=1)) +
  scale_y_continuous(limits=c(0,1))
```

# Bivariate Analysis

```{r}
#load the data 
osteoporosis <- read.csv2("osteoporosis.csv", sep = "\t", header = TRUE, dec = ",")
#see the data is correctly loaded
head(osteoporosis)
#overview of data
str(osteoporosis)
```

## Qualitative versus qualitative

* Numeric bivariate analysis

```{r}
#contingency table
table(osteoporosis$grupedad, osteoporosis$clasific)

#contingency table in %
prop.table(table(osteoporosis$grupedad, osteoporosis$clasific))
```

* Graphic analysis

```{r}
#plot the data: stacked barplot
ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific))

#plot the data II: bars side by side
ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific), position = "dodge")
```

```{r}
#Change colors, legend position, labels and finally save it!
p <- ggplot(data = osteoporosis, aes(x = grupedad)) +
  geom_bar(aes(fill = clasific), position = "dodge")
p + scale_fill_manual(values=c("#8618b1", "blanchedalmond", "red"))

p + theme(legend.position = "bottom")

p + labs(x = "Age group", y = "Women", title = "Osteo disease classified by age group")

pdf("clasific_grupedad.pdf")
  p + labs(x = "Age group", y = "Women", title = "Osteo disease classified by age group")
dev.off()
```

## Qualitative versus quantitative

* Numeric analysis

```{r}
#Table of statistics
with(osteoporosis, tapply(bua, list(grupedad), mean, na.rm=TRUE))
```

* Graphic analysis

```{r}
#Plot the data
bp <- ggplot(osteoporosis, aes(x = grupedad, y = bua)) + 
  geom_boxplot(fill = '#A4A4A4', color = "darkred")
bp

# Box plot with points
# 0.2 : degree of jitter in x direction
bp + geom_jitter(shape = 16, position = position_jitter(0.2)) +
   labs(x = "Age Group", y = "Women", title = "Osteo disease classified by age group")
```

###Exercise solution

**Study the relationship between menop and group of illness (clasific)**

```{r}
#explore variables
head(osteoporosis[,c("menop", "clasific")])
str(osteoporosis[,c("menop", "clasific")])

#Numeric summaries for two categorical variables: contigency table
table(osteoporosis$menop, osteoporosis$clasific)
addmargins(table(osteoporosis$menop, osteoporosis$clasific))
### proportions with respect to total
prop.table(table(osteoporosis$menop, osteoporosis$clasific))
###relative frequencies with respect to rows
prop.table(table(osteoporosis$menop, osteoporosis$clasific), margin=1)
###relative frequencies with respect to columns
prop.table(table(osteoporosis$menop, osteoporosis$clasific), margin=2)

#Graphic summaries for two categorical variables: barplot
ggplot(data = osteoporosis, aes(x = clasific)) +
  geom_bar(aes(fill = menop), position = "dodge")
ggplot(data = osteoporosis, aes(x = clasific)) +
  geom_bar(aes(fill = menop),position = "fill")
```

**Study if peso is different in each group of illness (clasific).**

```{r}
#explore variables
head(osteoporosis[,c("peso", "clasific")])
str(osteoporosis[,c("peso", "clasific")])

#Numeric summaries for one categorical and one continuous variables: table of statistics
with(osteoporosis, tapply(peso, list(clasific), mean, na.rm=TRUE))

#Graphic summaries for one categorical and one continuous variables: grouped boxplot
ggplot(osteoporosis, aes(x = clasific, y = peso)) + 
  geom_boxplot()
```

## Quantitative versus quantitative

```{r}
# Basic scatter plot
ggplot(osteoporosis, aes(x = peso, y = imc)) + 
  geom_point()

# Change the point size, and shape
ggplot(osteoporosis, aes(x = peso, y = imc)) +
  geom_point(size = 1, shape = 1)

# Color the points depending of another variable
ggplot(osteoporosis, aes(x = peso, y = imc, color = clasific, shape = clasific)) +
  geom_point()

#not always the correlation is good
ggplot(osteoporosis, aes(x = edad, y = imc)) + 
  geom_point()

#correlation matrix
pairs(osteoporosis[, c("edad", "peso",  "talla", "imc", "bua", "menarqui")])

#with ggplots
# install.packages("GGally")
library(GGally)
                 
ggpairs(osteoporosis, columns = c("edad", "peso",  "talla", "imc", "bua", "menarqui"), ggplot2::aes(colour = clasific))

```

# Correlation

Pearson correlation between imc and peso

```{r}
#Pearson correlation
cor(osteoporosis$imc, osteoporosis$peso, method = "pearson")
#the plot
ggplot(osteoporosis, aes(x = peso, y = imc)) + 
  geom_point() + 
  geom_smooth(method="lm")
```

Pearson correlation between bua and edad

```{r}
cor(osteoporosis$bua, osteoporosis$edad, method = "pearson")

#the plot
ggplot(osteoporosis, aes(x = edad, y = bua)) + 
  geom_point()
```

Spearman correlation between bua and edad

```{r}
#Spearman correlation
cor(osteoporosis$bua, osteoporosis$edad, method = "spearman")
```

