---
title: "Statistics with R. Testing with continuous variables"
author: "UEB"
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: show
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE)
```

```{r libraries , echo=FALSE, include=FALSE}
library(dplyr)
library(readxl)
library(magrittr)
library(ggplot2)
```

\newpage

# hta dataset

```{r hta}
hta <- read_excel("datasets/hta.xls")
hta[1:5,1:9]
```

# Checking Normality 

## Graphically

```{r ,eval=TRUE, mysize=TRUE, size='\\tiny'}
oldpar<-par(mfrow=c(1,1)) # Guarda los parámetros para el dibgujo
par(mfrow=c(2,2)) # Dibuja cuatro gráficos por grafico
with(hta, boxplot(tad1, main="Box-plot") )

with(hta, hist(tad1) )

with(hta, qqnorm(tad1, main="Normal QQplot") );with(hta, qqline(tas1) )

par(oldpar) # Vuelve a los parámetros de dibujo originales

```


## Normality Test

```{r normtest , mysize=TRUE, size='\\small'}
with(hta,shapiro.test(tad1) ) # Shapiro Wilk test

```



# One sample Test

```{r onetest , mysize=TRUE, size='\\small'}
with(hta,t.test(tad1,mu=90) ) # One sample T.test

```

# Two groups comparison in independent samples


```{r}
ggplot(hta, aes(x=grupo, y=tad1)) +
  geom_boxplot()
```


## Homogeneity of variances Test

Start with a descriptive analysis

```{r vartestDescript , mysize=TRUE, size='\\small'}
library(car)
hta%>% 
  group_by(grupo) %>% 
  summarise(var = sd(tad1))
```

Then do the test

```{r vartest , mysize=TRUE, size='\\small'}
with(hta,leveneTest(tad1~factor(sexo),center="median"))
```
  - p value is over 0.05
  - We can assume homogeneity of variances
  
  
  
## T test when variances are equal



```{r eqttest , mysize=TRUE, size='\\small'}
with(hta,t.test(tad1~factor(grupo),var.equal=TRUE ))

```
- Type I Error is over than 0.05
- We cannot reject mean equality



  
## T test when variances are unequal



```{r neqttest , mysize=TRUE, size='\\small'}
with(hta,t.test(tad1~factor(grupo),var.equal=FALSE ))

```
- Same conclusions as before
- Test is also known as Welch test



  
## U Mann-Whitney or Sum Rank non parametric test

```{r descriptiveComp , mysize=TRUE, size='\\small'}
hta%>% 
  group_by(sexo) %>% 
  summarise(median = median(tad1)) 
```


```{r umtest , mysize=TRUE, size='\\small'}
with(hta,wilcox.test(tad1~factor(grupo), 
                     alternative='two.sided',
                     exact=TRUE, correct=FALSE))
```

 - Null Hypothesis cannot be rejected
 
 
 
# Two groups comparison in dependent samples

## Paired T-test



```{r pairttest , mysize=TRUE, size='\\small'}
with(hta,t.test(tad1,tad12,paired=TRUE))

summary(hta$tad1)
summary(hta$tad12)
```
 - P value is over 0.05
 
    
## Paired Sign-Rank Wilcoxon Test
 
```{r pwilcest , mysize=TRUE, size='\\small'}
with(hta,wilcox.test(tad1,tad12,
     exact=TRUE, paired=TRUE))

```




# diabetes dataset

```{r , mysize=TRUE, size='\\tiny'}
diabetes <- read_excel("datasets/diabetes.xls")
diabetes[1:5,1:9]
sapply(diabetes, class)
diabetes_factor <- diabetes %>%
  mutate_if(sapply(diabetes, is.character), as.factor) %>%
  select (-numpacie)

```


# More than two groups comparison in independent samples

```{r}
diabetes%>% 
  group_by(ecg) %>% 
  summarise( n=n(),
    mean = mean(edat),
            sd=sd(edat)) 
```

```{r}
ggplot(diabetes, aes(x=ecg, y=edat)) +
  geom_boxplot()
```

## ANOVA

```{r}

anova<-aov(edat~ecg,data=diabetes_factor)
summary(anova)

```

### Tukey test for multiple comparison

```{r  , mysize=TRUE, size='\\tiny'}
# install.packages("multcomp")
library(multcomp)
tuk <- glht(anova, linfct = mcp(ecg = "Tukey"))

  print(summary(tuk)) # pairwise tests
  
```

**Confidence intervals:**

```{r  , mysize=TRUE, size='\\tiny'}  
print(confint(tuk, level=0.95)) # confidence intervals

```

**Multicomparison plot**

```{r  , mysize=TRUE, size='\\small', eval=TRUE}
par(mar=c(3,10,3,3))
plot(confint(tuk))
```

## Kruskal-Wallis Test (non-parametric)

```{r , mysize=TRUE, size='\\small'}
diabetes_factor%>% 
  group_by(ecg) %>% 
  summarise(median = median(edat)) 

kruskal.test(edat~ecg,data=diabetes_factor)
```

### Dunn Test for multiple comparison

```{r , mysize=TRUE, size='\\small'}
# install.packages("dunn.test")
library(dunn.test)
with(diabetes_factor,dunn.test(edat,ecg,method="bonferroni"))
```

\newpage

# Exercise solutions

## Exercise 1

a) Check the normality of tas1 variable in hta dataset

```{r}
oldpar<-par(mfrow=c(1,1)) # Guarda los parámetros para el dibgujo
par(mfrow=c(2,2)) # Dibuja cuatro gráficos por grafico
with(hta, boxplot(tas1, main="Box-plot") )

with(hta, hist(tas1) )

with(hta, qqnorm(tas1, main="Normal QQplot") );with(hta, qqline(tas1) )

par(oldpar) # Vuelve a los parámetros de dibujo originales
```


```{r}
with(hta,shapiro.test(tas1) ) # Shapiro Wilk test

```

b) Can it be accepted that the initial TAS is 120 in Hipertensive patients?

```{r}
with(hta,t.test(tas1,mu=120)) # One sample T.test
```

c) Find the 95% confidence interval for the mean of tas1 variable

From the test before we can find the confidence interval for the mean.

d) Extra: Can it be accepted that the initial TAS is higher than 120 in Hipertensive women?

```{r}
hta_w <- subset(hta, sexo=="MUJER")
with(hta_w,t.test(tas1,mu=120, alternative="greater")) # One sample T.test
```

## Exercise 2

Is TAD comparable at baseline time between groups?

We first plot the distribution of tad1 variables in each group

```{r}
ggplot(hta, aes(x=grupo, y=tad1)) +
  geom_boxplot()
```

Test for homogeneity of variances between the two groups


```{r}
hta%>% 
  group_by(grupo) %>% 
  summarise(var = sd(tad1)) 

with(hta,leveneTest(tad1~factor(grupo), center="median"))
```
  - p value is over 0.05
  - We can assume homogeneity of variances
  
We apply the t.test for equal variances

```{r}
with(hta,t.test(tad1~factor(grupo),var.equal=TRUE ))
```

- Differences between groups are not significant.

Using a non-parametric test:

```{r}
with(hta,wilcox.test(tad1~factor(grupo), alternative='two.sided', exact=TRUE, correct=FALSE))

hta%>% 
  group_by(grupo) %>% 
  summarise(median = median(tad1)) 
```

- We obtain similar results.

## Exercise 3

Is sistolic blood pressure (TAS) comparable between first and 12th measures?

We can use the paired t.test since sample size is big and 

```{r}
with(hta,t.test(tas1,tas12,paired=TRUE))

summary(hta$tas1)
summary(hta$tas12)
```
 - P value is below 0.05
 
    
With the non-parametric test Paired Sign-Rank Wilcoxon Test
 
```{r}
with(hta,wilcox.test(tas1,tas12,
     exact=TRUE, paired=TRUE))

```
 - We obtain similar results
 
## Exercise 4

a) Se efectúa un estudio de seguimiento a 1018 sujetos atendidos en una clínica de obesidad. Se mide el Indice de Masa Corporal(IMC) y el perfil lipídico. Al cabo de 12 meses se evalúa de nuevo el IMC y el colesterol estando interesados en cuantificar la
disminución de ambos parámetros.

Disponemos de dos variables cuantitativas medidas en dos momentos distintos (muestras apareadas). Se quiere analizar si
los valores pre y post sufren algun cambio, para ello debemos: en primer lugar evaluar la normalidad
de las variables. En caso de que se ajusten a una distribución normal, miramos la homegeneidad de
las variables y realizamos un test t de Student para muestras apareadas. En el caso de que no sigan
una distribución normal, realizaremos el test no paramétrico de Wilcoxon para muestras apareadas.

b) Se analizan un grupo de variables inmunológicas (leucocitos totales, linfocitos B, natural
Killer, etc) en una muestra de 102 hombres y 147 mujeres mayores de 65 años. Se está interesado en ver la existencia de diferencias por sexo.

Disponemos de distintas variables cuantitativas (leucocitos,linfocitos, etc. ) y una variable cualitativa, sexo. Queremos ver si existen diferencias de las variables cuantitativas según el sexo, para ello debemos: en primer lugar evaluar la normalidad de las variables. En caso de que se ajusten a una distribución normal, miramos la homegeneidad de las variables y realizamos un test t de
Student. En el caso de que no sigan una distribución normal, realizaremos el test no paramétrico U de Mann-Whitnney.

## Exercise 5

Are there differences between systolic pressure (sbp) and ECG in diabetic patients? 

Extra: And between the follow-up time (tempsviu) and the ECG? 

a) For each variable, perform some descriptives and check normality assumptions

* Check normality

```{r}
oldpar<-par(mfrow=c(1,1)) # Guarda los parámetros para el dibgujo
par(mfrow=c(2,2)) # Dibuja cuatro gráficos por grafico
with(diabetes, boxplot(sbp, main="Box-plot") )

with(diabetes, hist(sbp) )

with(diabetes, qqnorm(sbp, main="Normal QQplot") );with(diabetes, qqline(sbp) )

par(oldpar) # Vuelve a los parámetros de dibujo originales
```


```{r}
with(diabetes,shapiro.test(sbp) ) # Shapiro Wilk test

```

* Summarise variable by ECG group

```{r}
diabetes%>% 
  group_by(ecg) %>% 
  summarise( n=n(),
    mean = mean(sbp), median=median(sbp),
            sd=sd(sbp)) 
```

```{r}
ggplot(diabetes, aes(x=ecg, y=sbp)) +
  geom_boxplot()
```

b) Set the hypothesis contrast and perform the test

H0: The means of all the groups are equal
H1: Not all the means are equal. At least there are two different means

```{r}
anova_sbp <-aov(sbp~ecg,data=diabetes_factor)
summary(anova_sbp)

```

c) If needed, apply a post-hoc test for multiple comparisons between groups to test which groups are different.

Not needed here since we cannot reject the null hypothesis.

d) Extra: Analyze the relation between the follow-up time (tempsviu) and the ECG.

* Check normality of tempsviu

```{r}
oldpar<-par(mfrow=c(1,1)) # Guarda los parámetros para el dibgujo
par(mfrow=c(2,2)) # Dibuja cuatro gráficos por grafico
with(diabetes, boxplot(tempsviu, main="Box-plot") )

with(diabetes, hist(tempsviu) )

with(diabetes, qqnorm(tempsviu, main="Normal QQplot") );with(diabetes, qqline(tempsviu) )

par(oldpar) # Vuelve a los parámetros de dibujo originales
```


```{r}
with(diabetes,shapiro.test(tempsviu) ) # Shapiro Wilk test

```

* Summarise variable by ECG group

```{r}
diabetes%>% 
  group_by(ecg) %>% 
  summarise( n=n(),
    mean = mean(tempsviu), median=median(tempsviu),
            sd=sd(tempsviu)) 
```

```{r}
ggplot(diabetes, aes(x=ecg, y=tempsviu)) +
  geom_boxplot()
```

b) Set the hypothesis contrast and perform the test

H0: The means of all the groups are equal
H1: Not all the means are equal. At least there are two different means

We will perform a non-parametric test since we can't assume normality of the data

```{r}
diabetes_factor%>% 
  group_by(ecg) %>% 
  summarise(median = median(tempsviu)) 

kruskal.test(tempsviu~ecg,data=diabetes_factor)
```
- We can reject the null hypothesis (p-value < 0.05)

c) If needed, apply a post-hoc test for multiple comparisons between groups to test which groups are different.

Non-parametric Dunn Test for multiple comparison

```{r}
# install.packages("dunn.test")
library(dunn.test)
with(diabetes_factor,dunn.test(tempsviu,ecg,method="bonferroni"))
```
- Differences between Normal-Anormal groups and between Normal-Frontera groups are significant (adjusted p-value < 0.05).

## Exercise 6

Se estudia un grupo de 33 pacientes afectados de Carcinoma hepatocelular, un grupo de 22
afectados únicamente de cirrosis y un grupo control de 31 donantes de sangre. Se determina
la actividad celular NK y el número de células CK en los tres grupos. (Nota: la media de
actividad celular es de 39 unidades líticas /10ˆ7 de linfocitos y la mediana es de 28 y la media
del número de células es de 178 y la mediana de 163)

Solución: Se dispone de tres grupos de pacientes independientes y dos variables cuantitativas a comparar en los tres
grupos. Aunque desconocemos el comportamiento de las variables, por las pistas vemos que las medias y
las medianas son bastante diferentes con lo cual la distribución no debe de ser simétrica. Por otra parte el
tamaño muestral de los grupos no es muy grande. Optaríamos por efectuar un contraste no paramétrico de
igualdad de distribuciones en los tres grupos. Utilizaríamos la prueba de Kruskal-Wallis. Si hay diferencias,
para identificar entre qué dos grupos están usaríamos la corrección de Bonferroni del ejercicio anterior o el
Dunn-test que ya efectúa los contrastes con la p corregida.
